Containers and R---the Rockerverse and beyond

### 2020-01-16 manuscript.Rmd

Docker is perfectly suited to package and execute any software and data, and is a good candidate to build complex processing pipelines independent of the used programming language.
Because of the original use case (see&nbsp;\nameref{introduction}), Docker does not have a standard mechanism to chain a number of containers together, i.e., to define how environment variables, volume mounts, or ports can be used to pass input parameters and data into a container and how to get results out.
Some packages, e.g. \pkg{containerit}, provide a Docker image that can be used similar as a CLI, but the usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.

**\pkg{outsider}** ([https://antonellilab.github.io/outsider/](https://antonellilab.github.io/outsider/)) tackles the problem of integrating external programmes into an R workflow \citep{bennett_outsider_2019}.
Installation and usage of external programmes can be cumbersome and even prohibited if platform dependent.
Therefore \pkg{outsider} uses the platform-independent Docker images to encapsulate processes in _outsider modules_.
Each outsider modules has a `Dockerfile` and an R package with functions for interacting with the encapsulated tool.
A user only needs to use the given functions to install a module and can then integrate a tool into their own workflow seaminglessly using only R functions.
The module takes care of transmitting arguments and sending and receiving files to and from a container.
Outsider modules are hosted, e.g., on GitHub, and \pkg{outsider} contains discovery functions for them.
<!-- illustrating outsider examples:
https://github.com/DomBennett/om..astral/blob/master/examples/astral.R
https://github.com/DomBennett/om..raxml#partitioned-dna-analysis
-->

>>>>

Docker is perfectly suited for packaging and executing any software and data, and is a good candidate for building complex processing pipelines that are independent from the host programming language.
Due to its original use case (see&nbsp;\nameref{introduction}), Docker has no standard mechanisms for chaining containers together; it lacks definitions and protocols for environment variables, volume mounts and/or ports that could enable transfer of input (parameters and data) and output (results) to and from containers.
Some packages, e.g. \pkg{containerit}, do provide Docker images that can be used similarly to CLIs, but their usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.

**\pkg{outsider}** ([https://antonellilab.github.io/outsider/](https://antonellilab.github.io/outsider/)) tackles the problem of integrating external programs into an R workflow \citep{bennett_outsider_2020}.
Installation and usage of external programs can be difficult, convoluted and even impossible if the platform is incompatible. 
Therefore \pkg{outsider} uses the platform-independent Docker images to encapsulate processes in _outsider modules_.
Each outsider module has a `Dockerfile` and an R package with functions for interacting with the encapsulated tool.
Using only R functions, an end-user can install a module with the \pkg{outsider} package and then call module code to integrate a tool into their own workflow seamlessly.
The \pkg{outsider} package and module manage the containers and handle the transmission of arguments and the transfer of files to and from a container.
These functionalities also allow a user to launch module code on a remote machine via SSH, expanding the potential computational scale.
Outsider modules can be hosted code-sharing services, e.g., on GitHub, and \pkg{outsider} contains discovery functions for them.

<!-- @nuest, would you like me to expand on these in a particular way?
https://github.com/DomBennett/om..astral/blob/master/examples/astral.R
https://github.com/DomBennett/om..raxml#partitioned-dna-analysis
-->
